---
documentclass: book
classoption: oneside
title: "Exploring the utility of human genetic interactions for gene function predictions"
author: |
 | [Hai Fang](mailto:fh12355@rjh.com.cn)
date: "`r Sys.Date()`"
url: 'https\://23verse.github.io/fungi'
github-repo: "23verse/fungi"
description: "Function prediction by genetic interactions."
vignette: >
 %\VignetteIndexEntry{fungi Booklet}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document:
        toc: true
        toc_depth: 3
        number_sections: true
        theme: journal
        highlight: monochrome
        toc_float: true
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="##")
knitr::opts_chunk$set(dpi=300)
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(fig_cap="yes")
knitr::opts_chunk$set(eval=F)
```

```{r, eval=F, echo=F}
cd ~/Sites/Fang/FUNGI
################################################
R
library(tidyverse)
library(igraph)
library(Fang)

## load
load("now.FUNGI_booklet.RData")

placeholder <- "~/Sites/SVN/github/bigdata_dev/"
ig.BioGRID_genetic <- Fang::oRDS("ig.BioGRID_genetic", placeholder=placeholder)
org.Hs.egGOMF <- Fang::oRDS("org.Hs.egGOMF", placeholder=placeholder)

## save
res <- sapply(ls(), XGR::xObjSize, units="Mb") %>% sort()
items <- setdiff(ls(), c("img"))
items <- c('ig.BioGRID_genetic', 'org.Hs.egGOMF', 'df_pred', 'df_auc')
save(list=items, file="now.FUNGI_booklet.RData", compress="xz")

## bookdown::render_book + bookdown::gitbook
## use configuration in '_output.yml' and '_bookdown.yml'
bookdown::render_book("now.FUNGI_booklet.Rmd", c("bookdown::gitbook","bookdown::pdf_book","bookdown::tufte_html_book")[1])

q('no')

################################################

##########################
https://23verse.github.io/fungi
##########################
## bookdown::render_book + bookdown::gitbook
cd ~/Sites/SVN/github/fungi
# delete previous *.html *.png *.css *.js *.ttf *.Rmd files *.yml files
find ./ -type f -name '*.html' -o -type f -name '*.png' -o -type f -name '*.css' -o -type f -name '*.js' -o -type f -name '*.ttf' -o -type f -name '*.Rmd' -o -type f -name '*.yml' -o -type f -name '*.RDS' -o -type f -name '*.r' | xargs rm -rf
############
# copy ig.BioGRID_genetic.RDS and org.Hs.egGOMF.RDS to ~/Sites/SVN/github/fungi
cp -rf ~/Sites/SVN/github/bigdata_dev/ig.BioGRID_genetic.RDS ./
cp -rf ~/Sites/SVN/github/bigdata_dev/org.Hs.egGOMF.RDS ./
# copy oRDS.r and org.Hs.egGOMF.RDS to ~/Sites/SVN/github/fungi
cp -rf ~/Sites/XGR/oRDS.r ./
cp -rf ~/Sites/XGR/oRWkernel.r ./
cp -rf ~/Sites/XGR/xClassifyPerf.r ./
############
# copy the whole package from ~/Sites/Fang/FUNGI/FUNGIgitbook to ~/Sites/SVN/github/fungi
cp -rf ~/Sites/Fang/FUNGI/FUNGIgitbook/* ./
# copy/rename ~/Sites/Fang/FUNGI/now.FUNGI_booklet.Rmd to ~/Sites/SVN/github/fungi/fungi.Rmd
cp -rf ~/Sites/Fang/FUNGI/now.FUNGI_booklet.Rmd ./fungi.Rmd
cp -rf ~/Sites/Fang/FUNGI/_bookdown.yml ./_bookdown.yml
cp -rf ~/Sites/Fang/FUNGI/_output.yml ./_output.yml
cp -rf ~/Sites/Fang/FUNGI/now.cover.png ./cover.png
# add new *.html *.png *.css *.js *.ttf *.Rmd *.yml files
git status | grep 'modified' | awk '{print $2}' | xargs git add
git status | grep 'deleted' | awk '{print $2}' | xargs git rm
find ./ -type f -name '*.html' -o -type f -name '*.png' -o -type f -name '*.css' -o -type f -name '*.js' -o -type f -name '*.ttf' -o -type f -name '*.Rmd' -o -type f -name '*.yml' -o -type f -name '*.RDS' -o -type f -name '*.r' | xargs git add
git add --all
git commit -m 'Update fungi'
git push origin master
##########################

####################################################
## GitHub Pages

One site per account but unlimited projects can be created from https://pages.github.com:

> **One site `https://github.com/23verse/23verse.github.io`**

# visit https://github.com/new (enter "23verse.github.io" -> Initialize this repository with a README)
# clone into the local machine 
cd ~/Sites/SVN/github
git clone git@github.com:23verse/23verse.github.io
cd ~/Sites/SVN/github/23verse.github.io
git pull origin master
## https://github.com/23verse/23verse.github.io/settings/access: add "hfang-bristol"
echo "Hello 23verse" > index.html
git add --all
git commit -m 'Initial commit'
git push origin master

# Alive (https://23verse.github.io)

> **A project `https://github.com/23verse/fungi`**

# visit https://github.com/new (enter "fungi" -> Initialize this repository with a README)
## https://github.com/23verse/fungi/settings/access: add "hfang-bristol"
# clone into the local machine 
cd ~/Sites/SVN/github
git clone git@github.com:23verse/fungi.git
cd ~/Sites/SVN/github/fungi
git pull origin master
echo "Hello fungi" > index.html
git add --all
git commit -m 'Initial commit'
git push origin master
# visit https://github.com/23verse/fungi/settings -> GitHub Pages (master branch)

# Alive (https://23verse.github.io/fungi): Function predictions by genetic interactions
####################################################

```

# Overview {#index}

Human genetic interactions can be utilised in a variety of contexts, including computational predictions of gene functions using Gene Ontology (GO) terms. Leveraging the informativeness of such interactions however is not satisfactorily addressed. Here I introduce a new scoring metric `evaluation-prediction score` to meet this demand. This metric incorporates local network information building on short random walk. Applying this metric to the status quo genetic interactions in humans (also the physical interactions of the same source) reveals that although physical interactions are most informative based on direct neighbors (1-step), genetic interactions should be best leveraged considering 2-step network influential range. In this website, I will guide the users through step-by-step instructions on how to do so ([Get started]). I encourage the users to run through this before extending to their own applications.

# R packages & Datasets

> **R and packages**

At the time of writing, the most recent version of R is 4.0.2 (Dark and Stormy Night). The instructions of how to install R in different platforms (Linux, OS X and Windows) can be found at https://www.r-project.org where precompiled binaries are provided for download. For Linux users who do not have a `sudo` privilege, R could be installed from the source code at the home directory (that is, `$HOME`):

```{r, eval=F, echo=T}
wget https://cran.wu.ac.at/src/base/R-4/R-4.0.2.tar.gz
tar xvfz R-4.0.2.tar.gz
cd R-4.0.2
./configure --prefix=$HOME/R-4.0.2
make
make check
make install
$HOME/R-4.0.2/bin/R # start R
```

I highly recommend using a dedicated package `BiocManager` to install and update any packages that have been deposited into `Bioconductor` and `CRAN`, two repositories that are exclusive to each other so that a package cannot be deposited into both. `BiocManager` should be installed first in a conventional way (i.e. using the function `install.packages`), and then can be used to install other packages in a single step.

```{r, eval=F, echo=T}
# first, install the package BiocManager
install.packages("BiocManager")

# then install packages from Bioconductor and CRAN
BiocManager::install(c("igraph","tidyverse","patchwork","ggforce"), dependencies=T)
```

> **Gene annotations by GO Molecular Function terms**

GO annotations for human genes were sourced from NCBI (the gene2go.gz file). According to the true-path rule, these annotations were also propagated to ancestor terms. This dataset was preprocessed into an `SET` object, saved as an RDS-formatted file `org.Hs.egGOMF.RDS`, available at `https://github.com/23verse/fungi`). 

```{r, eval=T, echo=T}
org.Hs.egGOMF
```

> **Genetic interactions**

Human genetic interactions were obtained from BioGRID (version 3.5.185) [7], with ~3,200 genes (mapped to NCBI Entrez GeneID) and ~8,000 interactions (stored as an igraph object). This dataset was preprocessed into an `igraph` object (using the `igraph` package), saved as an RDS file `ig.BioGRID_genetic.RDS`, available at `https://github.com/23verse/fungi`). 

```{r, eval=T, echo=T}
ig.BioGRID_genetic
```

# Get started

> **Load packages and datasets**

```{r, eval=F, echo=T}
library(tidyverse)

## load functions
source("https://github.com/23verse/fungi/raw/master/oRDS.r")
source("https://github.com/23verse/fungi/raw/master/oRWkernel.r")
source("https://github.com/23verse/fungi/raw/master/xClassifyPerf.r")

## load datasets
placeholder <- "https://github.com/23verse/fungi/raw/master"
ig.BioGRID_genetic <- oRDS("ig.BioGRID_genetic", placeholder=placeholder)
org.Hs.egGOMF <- oRDS("org.Hs.egGOMF", placeholder=placeholder)
```


> **Do prediction**

```{r, eval=F, echo=T}
## pre-calculate affinity matrix estimated using 2-step short random walk kernel 
## on a network of genetic interactions (ig.BioGRID_genetic) 
df_mat <- oRWkernel(ig.BioGRID_genetic, steps=2) %>% as.matrix()

## define known gene annotations for each term
df_term <- org.Hs.egGOMF$info %>% mutate(ic=-log10(n/max(n))) %>% rename(term=name) %>% filter(n>=5)

## df_pred
## note: ~2 min runtime using single core (Mac)
message(sprintf("Start (%s) ...", as.character(Sys.time())), appendLF=T)
pb <- df_term %>% nrow() %>% progress_estimated()
df_term %>% select(term, member) %>% mutate(pred=map(member, function(x){
	pb$tick()$print()
	y <- df_mat
	ind <- which(!is.na(match(colnames(y),x)))
	## force to be 0 for the self of each seed gene
	diag(y)[ind] <- 0
	## how many genes used for the average
	vec <- rep(length(ind), nrow(y))
	vec[ind] <- length(ind) -1
	## calculate predictive score
	if(length(ind)>=5){
		res <- apply(y[,ind], 1, sum) / vec
		res %>% enframe(name="gene", value="score") %>% left_join(tibble(gene=x, seed=1), by='gene') %>% arrange(-score)
	}else{
		NULL
	}
})) %>% filter(map_lgl(pred,~!is.null(.x))) %>% select(term,pred) -> df_pred
message(sprintf("Finished (%s) ...", as.character(Sys.time())), appendLF=T)

## also append a list-column storing seed genes per term
df_pred <- df_pred %>% mutate(seed=map(pred, ~.x %>% filter(!is.na(seed)) %>% pull(gene)))
```

Gene-centric predictive score for each term is stored in a tibble `df_pred`, with 3 columns including 'term', 'pred' (a list-column containing details on gene-centric predictive score including 3 columns 'gene', 'score' for predictive score and 'seed' indicating whether or a seed gene), and 'seed' (seed genes).

```{r, eval=T, echo=T}
df_pred
```

> **Evaluate prediction performance**

```{r, eval=F, echo=T}
## df_auc
df_term_pred <- df_term %>% inner_join(df_pred, by='term')
pb <- df_term_pred %>% nrow() %>% progress_estimated()
df_term_pred %>% mutate(auc=map2_dbl(member, pred, function(x,y){
	pb$tick()$print()
	res <- xClassifyPerf(prediction=y, GSP=x, GSN=NULL, verbose=F)$auroc
})) -> df_auc
```

The tibble `df_auc` contains term-centric performance measured by AUC (the column 'auc').

```{r, eval=T, echo=T}
df_auc
```

Visualise term-specific AUC (see `A`), further stratified by information content of terms (see `B`).

```{r, eval=T, echo=T}
library(patchwork)
library(ggforce)

gp1 <- df_auc %>% ggplot(aes(x=namespace, y=auc, color=namespace)) + geom_violin() + ggforce::geom_sina(size=0.1, alpha=0.5) + theme_classic() + guides(color='none')
gp2 <- df_auc %>% ggplot(aes(x=ic, y=auc, color=namespace)) + geom_point(size=0.1) + theme_classic()  + guides(color='none')
(gp1 | gp2) + plot_annotation(tag_levels='A')
```


# Session Info

```{r sessionInfo, echo=F, eval=T}
library(magrittr)
library(dplyr)
XGR::xAuxSI() %>% dplyr::mutate(title=stringr::str_squish(title)) %>% select(c(2,3,4)) %>% knitr::kable(caption="Session info on R, OS, and attached/loaded packages", row.names=F)
#sessionInfo()
```