---
documentclass: book
classoption: oneside
title: "Exploring the utility of human genetic interactions for gene function predictions"
author: |
 | [Hai Fang](mailto:23versify@gmail.com)
 | 
 | University of Oxford 
url: 'https\://23verse.github.io/fungi'
github-repo: "23verse/fungi"
description: "Function prediction by genetic interactions."
vignette: >
 %\VignetteIndexEntry{fungi Booklet}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document:
        toc: true
        toc_depth: 3
        number_sections: true
        theme: journal
        highlight: monochrome
        toc_float: true
        code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="##")
knitr::opts_chunk$set(dpi=300)
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(fig_cap="yes")
knitr::opts_chunk$set(eval=F)
```

```{r, eval=FALSE, echo=FALSE}
cd ~/Sites/Fang/FUNGI
################################################
R

## load
load("now.FUNGI_booklet.RData")

placeholder <- "~/Sites/SVN/github/bigdata_dev/"
ig.BioGRID_genetic <- Fang::oRDS("ig.BioGRID_genetic", placeholder=placeholder)

## save
res <- sapply(ls(), xObjSize, units="Mb")
items <- setdiff(ls(), c("img","crosslink.customised"))
save(list=items, file="now.FUNGI_booklet.RData", compress="xz")

# https://bookdown.org/yihui/rmarkdown/html-document.html
# https://bootswatch.com/3/
###############
BiocManager::install(c("bookdown"))
###############

## bookdown::render_book + bookdown::gitbook
## use configuration in '_output.yml' and '_bookdown.yml'
bookdown::render_book("now.FUNGI_booklet.Rmd", c("bookdown::gitbook","bookdown::pdf_book","bookdown::tufte_html_book")[1])

q('no')

################################################

##########################
https://23verse.github.io/fungi
##########################
## bookdown::render_book + bookdown::gitbook
cd ~/Sites/SVN/github/fungi
# delete previous *.html *.png *.css *.js *.ttf *.Rmd files *.yml files
find ./ -type f -name '*.html' -o -type f -name '*.png' -o -type f -name '*.css' -o -type f -name '*.js' -o -type f -name '*.ttf' -o -type f -name '*.Rmd' -o -type f -name '*.yml' | xargs rm -rf
# copy the whole package from ~/Sites/Fang/FUNGI/FUNGIgitbook to ~/Sites/SVN/github/fungi
cp -rf ~/Sites/Fang/FUNGI/FUNGIgitbook/* ./
# copy/rename ~/Sites/Fang/FUNGI/now.FUNGI_booklet.Rmd to ~/Sites/SVN/github/fungi/fungi.Rmd
cp -rf ~/Sites/Fang/FUNGI/now.FUNGI_booklet.Rmd ./fungi.Rmd
cp -rf ~/Sites/Fang/FUNGI/_bookdown.yml ./_bookdown.yml
cp -rf ~/Sites/Fang/FUNGI/_output.yml ./_output.yml
cp -rf ~/Sites/Fang/FUNGI/now.cover.png ./cover.png
# add new *.html *.png *.css *.js *.ttf *.Rmd *.yml files
git status | grep 'modified' | awk '{print $2}' | xargs git add
git status | grep 'deleted' | awk '{print $2}' | xargs git rm
find ./ -type f -name '*.html' -o -type f -name '*.png' -o -type f -name '*.css' -o -type f -name '*.js' -o -type f -name '*.ttf' -o -type f -name '*.Rmd' -o -type f -name '*.yml' | xargs git add
git add --all
git commit -m 'Update fungi'
git push origin master
##########################

cd ~/Sites/XGR/MIMB/Submission/Accept
/System/Library/Automator/Combine\ PDF\ Pages.action/Contents/Resources/join.py -o fungi_Figures.pdf Figure1.ai Figure2.ai Figure3.ai Figure4.ai Figure5.ai Figure6.ai Figure7.ai Figure8.ai Figure9.ai Figure10.ai

####################################################
## GitHub Pages

One site per account but unlimited projects can be created from https://pages.github.com:

> **One site `https://github.com/23verse/23verse.github.io`**

# visit https://github.com/new (enter "23verse.github.io" -> Initialize this repository with a README)
# clone into the local machine 
cd ~/Sites/SVN/github
git clone git@github.com:23verse/23verse.github.io
cd ~/Sites/SVN/github/23verse.github.io
git pull origin master
## https://github.com/23verse/23verse.github.io/settings/access: add "hfang-bristol"
echo "Hello 23verse" > index.html
git add --all
git commit -m 'Initial commit'
git push origin master

# Alive (https://23verse.github.io)

> **A project `https://github.com/23verse/fungi`**

# visit https://github.com/new (enter "fungi" -> Initialize this repository with a README)
## https://github.com/23verse/fungi/settings/access: add "hfang-bristol"
# clone into the local machine 
cd ~/Sites/SVN/github
git clone git@github.com:23verse/fungi.git
cd ~/Sites/SVN/github/fungi
git pull origin master
echo "Hello fungi" > index.html
git add --all
git commit -m 'Initial commit'
git push origin master
# visit https://github.com/23verse/fungi/settings -> GitHub Pages (master branch)

# Alive (https://23verse.github.io/fungi): Function prediction by genetic interactions
####################################################

```

# Overview {#index}

As part of `Epistasis: Methods and Protocols`, [Methods in Molecular Biology - Springer](https://www.springer.com/gp/book/9781071609460), we describe three showcases analysing human genetic interactions and/or gene expression data in human tissues ([Materials]), compatible with the concept of tidy data science (FIGURE \@ref(fig:tidy) and FIGURE \@ref(fig:liner)). All done exclusively using the R one-liner, defined as a sequential pipeline of elementary functions chained together achieving a complex task. We will guide the users through step-by-step instructions on ([Case 1]) how to identify, visualise and interpret network modules of genetic interactions; ([Case 2]) how to identify and interpret tissue-specific genetic interactions; and ([Case 3]) how to carry out genetic interaction-based tissue clustering and differential interaction analysis. All three showcases are producible on its own, achieved in relatively short runtime (~20 min for Case 1, ~15 min for Case 2, and ~25 min for Case 3). We encourage the users to run through these showcases before analysing their own datasets.


# Materials

> **R and packages**

At the time of writing, the most recent version of R is 4.0.2 (Dark and Stormy Night). The instructions of how to install R in different platforms (Linux, OS X and Windows) can be found at https://www.r-project.org where precompiled binaries are provided for download. For Linux users who do not have a `sudo` privilege, R could be installed from the source code at the home directory (that is, `$HOME`):

```{r, echo=TRUE}
wget https://cran.wu.ac.at/src/base/R-4/R-4.0.2.tar.gz
tar xvfz R-4.0.2.tar.gz
cd R-4.0.2
./configure --prefix=$HOME/R-4.0.2
make
make check
make install
$HOME/R-4.0.2/bin/R # start R
```

We highly recommend using a dedicated package `BiocManager` to install and update any packages that have been deposited into `Bioconductor` and `CRAN`, two repositories that are exclusive to each other so that a package cannot be deposited into both. `BiocManager` should be installed first in a conventional way (i.e. using the function `install.packages`), and then can be used to install other packages in a single step. Once an additional package `remotes` also installed, `BiocManager` can be also used to install packages hosted at `GitHub`, usually as a development repository prior to submission into `Bioconductor` or `CRAN`.

```{r, echo=TRUE}
# first, install the package BiocManager
install.packages("BiocManager")

# then install packages from Bioconductor and CRAN
BiocManager::install(c("biobroom","dnet","ggrepel","gridExtra","limma","patchwork","remotes","tidyverse","osfr"), dependencies=T)

# can also install packages from GitHub
BiocManager::install("hfang-bristol/XGR")
```

> **Genetic interactions**

We extracted human genetic interactions from `BioGRID` (version 3.5.179) involving 3102 genes (mapped to NCBI GeneID; the same hereinafter) and their 7856 interactions. This dataset was preprocessed into an `igraph` object (using the `igraph` package), saved as an RData-formatted file `ig.BioGRID_genetic.RData`, deposited at `https://osf.io/gskpn`). 

```{r, eval=T, echo=T}
ig.BioGRID_genetic
```

# Session Info

```{r sessionInfo, echo=F, eval=T}
library(magrittr)
library(dplyr)
XGR::xAuxSI() %>% dplyr::mutate(title=stringr::str_squish(title)) %>% select(c(2,3,4)) %>% knitr::kable(caption="Session info on R, OS, and attached/loaded packages", row.names=F)
#sessionInfo()
```